{\bf \Large The Power of Context in Decision-Making and Recommendations}

### 2023-07-02 draft.tex
List of tables, list of figures, acknowledgements, and most formatting done.

\documentclass[a4paper,12pt]{article}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[lmargin=2.5cm,tmargin=3cm,rmargin=2.5cm,vscale=0.8,nohead]{geometry}
\usepackage{multicol}
\usepackage{wrapfig}
\usepackage{tikz}
\usepackage{tikz-network}
\usepackage{subfig}
\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{bibentry}
\usepackage[round]{natbib}
%\usepackage[numbers]{natbib}
\usepackage{xpatch}
\usepackage{pgfplots}
\newcommand{\citeyearonly}[1]{\citeyearpar{#1}}

\usepackage[activeacute,english]{babel}
%\usepackage{fancyhdr}
\usepackage{mathrsfs}
\usepackage{authblk}
\usepackage[alwaysadjust]{paralist}
\usepackage{alltt}
\usepackage{caption}
\usepackage{array}
\usepackage[fit]{truncate}
%\pagestyle{fancy}
\usepackage{calc}
%\usepackage{fancyvrb}
\usepackage{float}
\usepackage[ansinew]{inputenc}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{tabularx}
\usepackage[figuresright]{rotating}
\usepackage{color}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usetikzlibrary{positioning, arrows}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{array}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{array}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{adjustbox}
\usepackage[stable]{footmisc}

@@

@@

\title{{\bf \Large More than Meets the Eye: The Power of Context in Decision-Making and Recommendations}
}

@@

\begin{document}
\spacing{1.5}
\maketitle

@@

\end{abstract}
\newpage
\section{Introduction}

@@

Choice context has long received the attention of scholars in many fields, including marketing, psychology, economics. Because it was studied in many domains, scholars refer to it differently. This thesis concentrates and uses two of them. The first one posits that choice context \footnote{For the remainder of the dissertation, I will follow the previous literature \citep{truebloodEtAl13} and refer this definition of choice context as context effects.} is ``the availability and the nature of choice alternatives" \citep{tversky1972elimination, huberEtAl82, simonson89}. Previous research has demonstrated the existence of context effects in various settings \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  wuConsguner20}. The second definition of context arises much later with the proliferation of recommender systems. Context also referred to as ``the time and content of the choice, the location or socio-demographic characteristics of the decision-maker..." \citep{adomavicius2011context}\footnote{To distinguish between these two defintions I will refer to this definition as external context throughout this dissertation.}

However, modeling for these effects mathematically was a challenge because the existing models suffered from independence of irrelevant alternatives \citep{luce59}, meaning these models treated each alternative in isolation. This has shifted the scholar's interest towards computational models \citep{usher2001time, roe2001multialternative, trueblood2014multiattribute, noguchi2018multialternative}. Yet, none of these models have been to real-world choice data and as a result, their applicability to field data remained an open question. I address this gap in the first study where I apply a computational model to a field data of high heterogeneity among dimensions. Building on my results in my second study I delve into the choice modeling more and develop a methodology to account for three main context effects studied in the literature \citep{truebloodEtAl13}.

When new users or items are introduced to recommender systems, they fail to function as intended because of the lack of information their algorithms need. In such cases, even when they do provide recommendations, they are far from personalized \citep{lika2014facing}. It is considered to be a key challenge in recommender system design \citep{park2009pairwise}. Previous literature has applied various methods to overcome this problem \citep{guy2009personalized, aharon2013off, bykau2013coping, saveski2014item}. However, the limitations of these approaches are that they ignore the context effects and concentrate on scenarios where information scarcity is temporary. Using the findings from my second study, decision-making and choice modeling literature I propose an innovative solution to address continious cold-start issue in recommender system design by utilizing menu contextual information to generate relevant choice sets through a two-step choice modeling method.

Research agrees on importance of metrics beyond accuracy for recommender systems \citep{kaminskas2016diversity}. It is in the best interest of online marketplaces to provide users with recommendations which are not only in the category ``exactly what I want", but also ``I never thought I would have liked this one" \citep{kotkovSurveySerendipityRecommender2016}. This can not only boost sales \citep{songWhenHowDiversify2019}, but also increase the satisfaction of users \citep{knijnenburgExplainingUserExperience2012}. Albeit, it was observed that the preferences of individuals are not stable and tend to change and system designers have proposed various tools to enable users signal their preference shifts \citep{bostandjiev2012tasteweights,hijikata2012relation}. However, to better understand it is necessary to study those tools in isolation from the context arising from the choice set and concentrate purely on external context that arises from the user side, e.g their preferences \citep{adomavicius2011context}. In my last study I address this gap by conducting an online experiment and applying Technology Acceptance Model \citep{davis1985technology} to measure users' acceptance of such tools.

@@

\section{Beyond the Experiment: Exploring Context Effects in Multi-Attribute, Multi-Alternative Choice Environments}\label{chapter:simulationStudy}
\begin{abstract}

@@

Also, applying MDbS to observational data would allow me to statistically asses the significance of the contribution of this model's ability to account for context effect. Instead of testing this model against established choice models, I will attempt to complement them with MDbS. To do this, random effect Probit model as a variation of Probit family will be used and augmented with MDbS output. Thee Probit model is chosen as it does not explictly assume IIA unlike family of logit models. To validate my results further I will apply the same methodology and analysis to an experimental data.

@@

If this situation is familiar to you, then you have experienced so called context effect which can bu understood as ``the composition  and the nature of the choice set, availability of various options in it" \citep{tversky1972elimination, huberPuto83}. For the long time, our understanding of the choice did not extend over the borders of two related principles. First one is, independence of irrelevant alternatives (IIA) which states that when having a choice between two two options $A$ and $B$ if a person prefers $A$ for example, regardless of adding a third option $C$ to this choice set, that person's preference must be unaltered \citep{luce59}. Second one is regularity principle which states that, the choice probability of option $A$ can not increase by introduction of option $C$ \citep{luce59}.

However, research has concluded that the context of the choice set and options in it have substantial effect on how people make choices. This effect has been studied extensively for the past five decades by many economists, marketing scholars and psychologists \footnote{See Dowling et al. \citeyearonly{dowlingEtAl20} and Lichtenstein \citeyearonly{lichtenstein2006construction} for more comprehensive review.} \citep{ kahnemanTversky79, simonson89, tverskySimonson93, lichtenstein2006construction, dowlingEtAl20}. Most of the research has focused on three context effects, also reffered to as ``big three": attraction, compromise and similarity \citep{howes2016contextual}.

@@

Now, lets add a third option to this choice set, option $D_A$ to create one variation and $D_B$ to create the second variation of the ternary choice set. Both added options have lower values in both dimensions when compared to $A$ and $B$ respectively. Huber created this type of scenario and has found what he has then called ``attraction effect" \citeyearonly{huberEtAl82}. Attraction effect, which is also known in the literature as asymmetric dominance effect is a consistent violation of the regularity principle mentioned earlier. He suggested that when having a choice set consisting of options $A$ and $B$ the relative probability of choosing option $A$ can be increased if one adds a third option with characteristics of $D_A$ to the same choice set \citeyearonly{huberEtAl82}.

@@

It is worth noting that, it is also possible to ``target" a particular option from binary choice set when adding a third option to create a compromise effect. One can add a target $C$ which makes $A$ a compromise option. In this case, the probability of choosing $A$ among this triple will increase, as it is considered a compromise between the remaining two options.

@@

Although Becker et al. \citeyearonly{becker1964measuring} have mentioned it before, the first study of similarity effect is known to be the one by Tversky \citep{tversky1972elimination}. He noted that when facing a binary choice set consisting of $A$ and $B$ individuals will gravitate towards $A$ more than when facing a ternary choice set consisting of ${A, B, S_A}$ depicted in figure \ref{fig:similarityEffect}. He explained it by proposing elimination by aspects theory, which states that one attribute will be chosen as elimination criteria, and all options that do not meet that criteria will be eliminated \citep{tversky1972elimination}. Hence in the choice set ${A, B, S_A}$ if an individual selects dimension 2 as elimination criteria, both $A$ and $S_A$ will be eliminated, leaving $B$ as a choice. On the contrary, if decision maker prefers dimension 1 more, then option 
$B$ will be eliminated, leaving both $A$ and $S_A$ to share the ``victory", hence resulting in equation \ref{eq:similarityProbability}.

@@

MDFT has been confirmed to account for similarity, compromise, and attraction effects in multialternative choice scenarios \citep{roe2001multialternative}. It has previously been tested against such random utility models of choice as logit and Probit and has been concluded to be a better fit to empirical data \citep{berkowitsch2014rigorously}. 
MDFT has been further adapted to account for preference shifts \citep{mohr2017attraction} and decision making under time restrictions \citep{diederich2003mdft}.

@@

According to MDbS, when comparing two tickets between Paris and New-York, the price of a ticket would be compared to the prices of other tickets in the choice set and also to the ones and which an individual has previously seen, but are not in the current choice set. Comparisons are ordinal, meaning that evidence accumulated towards the ``winner" at a rate of one irregardless of how large the difference was.

@@

\begin{align}\label{probabilityOneIsFavored}
    P(m_{ij} \text{ is favored over } m_{kj}) = 
        \begin{cases}
        F(\beta_0 (| \frac{m_{ij} - m_{kj}}{m_{kj}} |- \beta_1)) & \text{if } A_i > X_i \\
        0 & \text{otherwise}
        \end{cases} ,
\end{align}
where $F$ is a logistic sigmoid function and $\beta_0$ and $\beta_1$ corresponde to the advantage value and the probability that this particular advantage value will be enough to be preferred. For example, consider the case where $\beta_0 = 0.1$ and $\beta_1 = 50$. This would mean that the advantage of 10\% would be preferred with 50\% probability. Consequently, if the difference is 20\%, then it will be preferred with 99\% probability. The logistic function brings the notion of ``soft" comparison instead of ``hard" comparison in which case the small differences would be ignored, while large differences would be extremely preferred \citep{noguchi2018multialternative}.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{staticFiles/contextEffectExampleScatterplot.png}
    \caption{Example choice set to explain MDbS' account for big three context effects. $A$ and $B$ are considered original two alternatives (binary choice set). $D$ is dominated by $A$ on both dimensions. $C$ acts as compromise between the original two and $S$ is a similar option to $B$. Although these effects would be present in different variations of the choice set (for example one can make $A$ as compromise option), for the simplicity I will concentrate on this example.} % Add your description here
    \label{fig:MDBsContextExample} % This labels your figure for reference

@@

When adding option $D$ which is dominated by $A$ in both dimensions to binary choice set $A$ $B$ (hereafter binary choice set will be used instead of $A$ and $B$) one creates an attraction effect \citep{huberEtAl82, huberPuto83}. Huber et al. \citeyearonly{huberEtAl82} explained attraction effect via weight shifts for individuals. Addition of $D$ would make people to weigh dimension 2 more. Hence, $A$ and $D$ will have higher ``interest" among people, and this is where $A$ wins over $D$.

@@

In MDbS this is explained by $\beta_0$ and $\beta_1$ from equation \ref{probabilityOneIsFavored}. Recall that, the reason of having the sigmoid function with arguments $\beta_0$ and $\beta_1$ is to make sure that small differences would be relatively ignored. This is in line with previous literature stating that people tend to ignore small differences between alternatives when making a choice \citep{kalwani1992consumer}. Hence, the small differences between $B$ and $S$ would be ignored, which would be translated to decreased probability of any of them winning over the other when being compared. This will indirectly increase the evidence accumulation for $A$ while resulting in ``shared" evidence between $B$ and $S$.

@@

Because of MDbS nature of comparing dimensions with one another, for the purposes of this study, I can not use variables that do not differ across the menus. I can not utilize horizontal attributes in my analysis because they do not follow the standard ``greater the better" mathematical approach to comparison either. However, these variables are used later in this dissertation in chapters \ref{chapter:jmrPaper} and \ref{chapter:hicssPaper}. Hence, I am bound to utilize solely four vertical attributes in my analysis. Table \ref{tab:descriptiveStats} provides descriptive information about variables. I have also multiplied all four vertical variables by \-1 to convert them to negative scale because of MDbS nature of comparing absolute values and the assumption that consumers prefer lower values along vertical dimensions.

@@

Choice set designers and engineers have long used ``Top n" accuracy metrics when designing choice sets, or testing the performance of statistical models \citep{ricci2015recommender}. ``Top $n$" accuracy metric posits whether or not, true class of the option matches the top $n$ predictions of the model. I will follow and adopt this metric because it is well established in the literature, mirrors the real-world decision making and it fits the contribution of the thesis the best. I will use ``Top 1" accuracy which ranges from 0 to 1 as my metric of optimization for an individual choice set. Because choice sets in the data vary significantly in size, I will use average Top 1 accuracy metric weighted by menu size.This will ensure that smaller menus contribute proportional to their sizes. Also, to comply with differential evolution algorithm's aim of minimization, I will multiply this metric by -1. As an additional measure to explore the parameter space thoroughly, I employ multiple-run approach for differential evolution algorithm. Specifically, I will execute it ten times across the entire dataset. This reposition will allow me to further explore the parameter space, mitigating the risk of missing any region that can potentially contain optimal solution.

>>>>

\documentclass[a4paper,12pt]{article}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[lmargin=2.5cm,tmargin=3cm,rmargin=2.5cm,vscale=0.8,nohead]{geometry}
\usepackage{multicol}
\usepackage{wrapfig}
\usepackage{tikz}
\usepackage{tikz-network}
\usepackage{subfig}
\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{bibentry}
\usepackage[round]{natbib}
%\usepackage[numbers]{natbib}
\usepackage{xpatch}
\usepackage{pgfplots}
\newcommand{\citeyearonly}[1]{\citeyearpar{#1}}
%\usepackage{fancyhdr}
\usepackage{mathrsfs}
\usepackage{authblk}
\usepackage[alwaysadjust]{paralist}
\usepackage{alltt}
\usepackage{caption}
\usepackage{array}
\usepackage[fit]{truncate}
%\pagestyle{fancy}
\usepackage{calc}
%\usepackage{fancyvrb}
\usepackage{float}
\usepackage[utf8]{inputenc}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{tabularx}
\usepackage[figuresright]{rotating}
\usepackage{color}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usetikzlibrary{positioning, arrows}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{array}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{array}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{adjustbox}
\usepackage[stable]{footmisc}
\usepackage[french, english]{babel}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{epigraph}
\usepackage[page,toc,titletoc,title]{appendix}
\pgfplotsset{compat=1.18}

@@

\newenvironment{dedication}
  {%\clearpage           % we want a new page          %% I commented this
   \thispagestyle{empty}% no header and footer
   \vspace*{\stretch{1}}% some space at the top
   \itshape             % the text is in italics
   \raggedleft          % flush to the right margin
  }
  {\par % end the paragraph
   \vspace{\stretch{3}} % space at bottom is three times that at the top
   \clearpage           % finish off the page
  }

@@

\title{{\bf \Large The Power of Context in Decision-Making and Recommendations}
}

@@

\begin{document}

\begin{dedication}

To my father, my first mentor, who seeded my enduring passion for learning.

\end{dedication}

\spacing{1.5}
\maketitle

\clearpage
\begin{otherlanguage}{french}
\begin{abstract}

    Cette dissertation comprend quatre études qui étudient les effets du contexte et appliquent ces connaissances pour améliorer les systèmes de recommandation dans les marchés en ligne. Je distingue deux types de contexte. Le premier type est étroitement lié aux alternatives présentes dans un ensemble de choix. Le deuxième type de contexte provient de facteurs externes à l'ensemble de choix. Les conclusions de la dissertation indiquent que les effets de contexte internes dans les situations de choix, qui étaient principalement étudiés et observés dans des cadres expérimentaux auparavant, sont également présents et détectables dans des cadres réels. De plus, l'étude portant sur le contexte externe trouve une relation positive entre les outils qui permettent aux utilisateurs de signaler leurs préférences aux systèmes de recommandation et leur adoption. Les applications empiriques des chapitres qui composent cette thèse reposent sur quatre ensembles de données distincts et uniques, le plus grand étant des données observationnelles provenant d'un cadre réel et les trois autres provenant d'un cadre expérimental.

    Le premier chapitre applique un modèle de prise de décision computationnel à un ensemble de données de choix réel conséquent, ce qui constitue la première application de cette envergure. Les résultats indiquent que le contexte influence les choix des individus. L'étude suggère que les marchés en ligne pourraient utiliser de tels modèles pour approfondir leur compréhension de la manière dont la composition de l'ensemble de choix et l'interaction entre les différentes options affectent les décisions des consommateurs. La caractéristique clé de ce modèle computationnel est qu'il identifie le contexte de manière globale. Comprendre les composants sous-jacents du contexte peut favoriser une compréhension plus profonde, offrant des perspectives plus granulaires sur la manière dont chacun d'eux influence individuellement les comportements de choix des individus lorsqu'ils coexistent avec d'autres effets.

    En s'appuyant sur les résultats du premier chapitre, le deuxième chapitre va au-delà des études traditionnelles de contexte, en développant une méthodologie qui vise à dissocier ses trois principaux composants connus sous le nom d'attraction, de compromis et de similarité les uns des autres. Cette méthodologie est conçue pour être polyvalente, capable de gérer diverses situations de choix de tailles variées. Les données provenant à la fois des environnements en ligne et hors ligne peuvent être intégrées avec succès dans ce cadre, en élargissant sa portée et son applicabilité. Les perspectives dérivées de cette étude offrent un potentiel pour affiner les offres de produits en ligne et les systèmes de recommandation. Une application notable de cette approche réside dans la navigation du problème du 'démarrage à froid' auquel sont confrontés les marchés numériques. Elle préconise la conception d'ensembles de choix qui contiennent des alternatives empiriquement similaires, favorisant une prise de décision efficace dans des espaces numériques encombrés. Cette approche innovante représente une première étape pour la recherche future visant à extraire et à utiliser des informations contextuelles à partir des ensembles de choix. Par conséquent, elle contribue au développement des conceptions de systèmes de recommandation et à une compréhension plus approfondie de la nature hétérogène de la dynamique des choix des consommateurs.

    La troisième étude présente une nouvelle approche pour résoudre le problème du 'démarrage à froid' côté utilisateur dans la conception des systèmes de recommandation. Elle applique les résultats du processus de choix en deux étapes observé chez les individus aux informations contextuelles extraites de l'ensemble de choix par le biais du clustering pour générer des ensembles de choix. L'approche s'avère robuste à la méthode de clustering et aux différentes heuristiques appliquées. Ses conclusions ouvrent la voie à l'étude des effets de contexte provenant de l'extérieur des ensembles de choix, à savoir les préférences des individus et les outils qui leur permettent de signaler leurs préférences aux systèmes de recommandation. Il est soutenu que c'est l'un des principaux mécanismes pour créer des systèmes de recommandation plus efficaces.

    La dernière étude a examiné l'effet du contrôle de l'utilisateur sur l'acceptation du système de recommandation en utilisant le modèle d'acceptation de la technologie comme cadre théorique. Cette étude a constaté que les systèmes de recommandation faciles à utiliser étaient perçus comme plus utiles par les utilisateurs et entraînaient une plus grande intention de les utiliser. Cependant, différents mécanismes de contrôle avaient des impacts variés sur l'expérience de l'utilisateur. Les résultats proposent une opportunité passionnante pour la recherche future dans la conception de systèmes de recommandation plus efficaces, y compris l'intégration potentielle de grands modèles de langage (LLM) pour une meilleure interaction avec l'utilisateur et une transparence des recommandations.

    Cette thèse démontre l'existence d'effets de contexte dans des paramètres multi-attributs, multidimensionnels et développe des méthodologies pour améliorer la conception des systèmes de recommandation avec des effets de contexte. De plus, cette étude examine le contexte externe, à savoir, les outils qui permettent aux utilisateurs d'exprimer leurs préférences et comment améliorer les systèmes de recommandation grâce à eux.

\end{abstract}
\end{otherlanguage}

\clearpage
\begin{abstract}

@@

\end{abstract}
\clearpage

\section*{Acknowledgements}

Embarking on this scholarly journey has been both humbling and transformative, marked by countless days of meticulous study, intense deliberation, and profound discoveries. Indeed, the journey was punctuated with hurdles and moments of self-doubt. However, every trough was succeeded by a crest, each stumble by an ascent, and every challenge served as an opportunity. The path was steep, but it was these uphill struggles that honed my perspective, expanded my horizons, and ignited my passion for knowledge. The lessons learned, the triumphs celebrated, and the sense of accomplishment far outweighed the trials faced. Reflecting on these uplifting moments, I joyfully acknowledge those who accompanied me on this journey and made it not only possible but immensely rewarding.

Firstly, I extend my deepest gratitude to Zakaria Babutsidze, my PhD supervisor. His pledge to transition my perspective from engineering to social science laid the foundation for this journey. I am indebted to Francesco Castellanetta, our PhD program director, for his guiding hand and unwavering faith in my capabilities. To Ludovic Diabaggio, I appreciate our intriguing discussions that transcended academia. I am grateful to William Rand for his invaluable insights and co-authorship. Daniela Iubatti, your informal talks and unwavering support have been a source of inspiration. I appreciate Diego Zunino for his earnest and invaluable guidance. Bruno Cirillo deserves special mention; our albeit slightly formal discussions have significantly shaped my work. My heartfelt appreciation goes to Renata Kaminska. Her support and intuitive understanding of my thoughts were indispensable in my pursuits. I am grateful to Benjamin Montmartin; our separate exchanges about wine (we were in France in the end) and statistics enriched my perspectives immeasurably. To Valerio Incerti, our creative discourses, even about less creative ideas were a constant source of inspiration. Lapo Mola's invaluable input became a source of aspiration, for which I am grateful.

Aytan, my beloved wife and best friend, your unwavering support and guidance have been my beacon during tumultuous times. Ecem Delicik, my dear friend, I am grateful for the shared stories, laughter, and silence that have strengthened our friendship. The moment you realized I was speaking to you in Turkish is unforgettable. Johanna Deperi, our shared moments and long talks were invaluable for self-reflection, joy, and sometimes, ``justified sadness'' throughout this journey. Artyom Yepremyan, our discussions helped us conquer the PhD mountain. I will forever cherish the times when I needed a hand, and you extended yours. Manon Desjardins, your constant support despite the towering stack of papers on your desk will always be appreciated. Mehdi Ibrahim, I am grateful for your steady guidance and the irresistible desserts from Morocco. Teymur Mardaliyev, your friendship and coding prowess have been invaluable; only you could convince me to use JavaScript there.

Finally, no amount of words can fully express my gratitude to my parents. Dad, your inquisitive nature has been my guiding star, instilling in me the curiosity that set me on this scholarly path. Your enthusiasm for the uncharted has been infectious, leading me to question, explore, and discover. Mom, you have been my anchor, reminding me of my roots and my identity amidst the academic whirlwind. Your unwavering encouragement and steadfast belief have reinforced my sense of self, allowing me to face challenges with courage and resilience. You taught me to stay grounded and true to myself, no matter how far I travel in the pursuit of knowledge. Your lessons continue to shape my journey, helping me navigate both personal and academic hurdles with grace and fortitude. Çox sağ olun ki, varsınız!

\clearpage
\tableofcontents
\newpage
\listoffigures
\clearpage
\listoftables
\newpage

\newpage
\section{Introduction}

\epigraph{We all make choices, but in the end, our choices make us.}{}

@@

Choice context has long received the attention of scholars in many fields, including marketing, psychology, economics. Because it was studied in many domains, scholars refer to it differently. This thesis concentrates and uses two of them. The first one posits that choice context \footnote{For the remainder of the dissertation, I will follow the previous literature \citep{truebloodEtAl13} and refer to this definition of choice context as context effects.} is ``the availability and the nature of choice alternatives'' \citep{tversky1972elimination, huberEtAl82, simonson89}. Previous research has demonstrated the existence of context effects in various settings \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  wuConsguner20}. The second definition of context arises much later with the proliferation of recommender systems. Context also referred to as ``the time and content of the choice, the location or socio-demographic characteristics of the decision-maker...'' \citep{adomavicius2011context}\footnote{To distinguish between these two definitions I will refer to this definition as external context throughout this dissertation.}

However, modeling for these effects mathematically was a challenge because the existing models suffered from independence of irrelevant alternatives criterion \citep{luce59}, meaning these models treated each alternative in isolation. This has shifted the scholar's interest towards computational models \citep{usher2001time, roe2001multialternative, trueblood2014multiattribute, noguchi2018multialternative}. Yet, none of these models has been applied to real-world choice data and as a result, their applicability to field data remained an open question. I address this gap in the first study where I apply a computational model to a field data of high heterogeneity among dimensions. Building on my results in my second study I delve into the choice modeling more and develop a methodology to account for three main context effects studied in the literature \citep{truebloodEtAl13}.

When new users or items are introduced to recommender systems, they fail to function as intended because of the lack of information their algorithms need. In such cases, even when they do provide recommendations, recommendations are far from personalized \citep{lika2014facing}. It is considered to be a key challenge in recommender system design \citep{park2009pairwise}. Previous literature has applied various methods to overcome this problem \citep{guy2009personalized, aharon2013off, bykau2013coping, saveski2014item}. However, the limitations of these approaches are that they ignore the context effects and concentrate on scenarios where information scarcity is temporary. Using the findings from my second study, decision-making and choice modeling literature I propose an innovative solution to address continuous cold-start issue in recommender system design by utilizing contextual information of the menu to generate relevant choice sets through a two-step choice modeling method.

Research agrees on importance of metrics outside the boundaries of accuracy for recommender systems \citep{kaminskas2016diversity}. It is in the best interest of online marketplaces to provide users with recommendations which are not only in the category ``exactly what I want'', but also ``I never thought I would have liked this one'' \citep{kotkovSurveySerendipityRecommender2016}. This can not only boost sales \citep{songWhenHowDiversify2019}, but also increase the satisfaction of users \citep{knijnenburgExplainingUserExperience2012}. Albeit, it was observed that the preferences of individuals are not stable and tend to change and system designers have proposed various tools to enable users signal their preference shifts \citep{bostandjiev2012tasteweights,hijikata2012relation}. However, to better understand it is necessary to study those tools in isolation from the context arising from the choice set and concentrate purely on external context that arises from the user side, e.g their preferences \citep{adomavicius2011context}. In my last study I address this gap by conducting an online experiment and applying Technology Acceptance Model \citep{davis1985technology} to measure users' acceptance of such tools.

@@

\section{Exploring Context Effects in Multi-Attribute, Multi-Alternative Choice Environments}\label{chapter:simulationStudy}
\begin{abstract}

@@

Also, applying MDbS to observational data would allow me to statistically asses the significance of the contribution of this model's ability to account for context effect. Instead of testing this model against established choice models, I will attempt to complement them with MDbS. To do this, random effect Probit model as a variation of Probit family will be used and will be augmented with MDbS output. The Probit model is chosen as it does not explicitly assume IIA unlike family of Logit models. To validate my results further I will apply the same methodology and analysis to an experimental data.

@@

If this situation is familiar to you, then you have experienced so called context effect which can bu understood as ``the composition  and the nature of the choice set, availability of various options in it'' \citep{tversky1972elimination, huberPuto83}. For the long time, our understanding of the choice did not extend over the borders of two related principles. First one is, independence of irrelevant alternatives (IIA) which states that when having a choice between two two options $A$ and $B$ if a person prefers $A$ for example, regardless of adding a third option $C$ to this choice set, that person's preference must be unaltered \citep{luce59}. Second one is regularity principle which states that, the choice probability of option $A$ can not increase by introduction of option $C$ \citep{luce59}.

However, research has concluded that the context of the choice set and options in it have substantial effect on how people make choices. This effect has been studied extensively for the past five decades by many economists, marketing scholars and psychologists \footnote{See Dowling et al. \citeyearonly{dowlingEtAl20} and Lichtenstein \citeyearonly{lichtenstein2006construction} for more comprehensive review.} \citep{ kahnemanTversky79, simonson89, tverskySimonson93, lichtenstein2006construction, dowlingEtAl20}. Most of the research has focused on three context effects, also reffered to as ``big three'': attraction, compromise and similarity \citep{howes2016contextual}.

@@

Now, lets add a third option to this choice set, option $D_A$ to create one variation and $D_B$ to create the second variation of the ternary choice set. Both added options have lower values in both dimensions when compared to $A$ and $B$ respectively. Huber created this type of scenario and has found what he has then called ``attraction effect'' \citeyearonly{huberEtAl82}. Attraction effect, which is also known in the literature as asymmetric dominance effect is a consistent violation of the regularity principle mentioned earlier. He suggested that when having a choice set consisting of options $A$ and $B$ the relative probability of choosing option $A$ can be increased if one adds a third option with characteristics of $D_A$ to the same choice set \citeyearonly{huberEtAl82}.

@@

It is worth noting that, it is also possible to ``target'' a particular option from binary choice set when adding a third option to create a compromise effect. One can add a target $C$ which makes $A$ a compromise option. In this case, the probability of choosing $A$ among this triple will increase, as it is considered a compromise between the remaining two options.

@@

Although Becker et al. \citeyearonly{becker1964measuring} have mentioned it before, the first study of similarity effect is known to be the one by Tversky \citep{tversky1972elimination}. He noted that when facing a binary choice set consisting of $A$ and $B$ individuals will gravitate towards $A$ more than when facing a ternary choice set consisting of ${A, B, S_A}$ depicted in figure \ref{fig:similarityEffect}. He explained it by proposing elimination by aspects theory, which states that one attribute will be chosen as elimination criteria, and all options that do not meet that criteria will be eliminated \citep{tversky1972elimination}. Hence in the choice set ${A, B, S_A}$ if an individual selects dimension 2 as elimination criteria, both $A$ and $S_A$ will be eliminated, leaving $B$ as a choice. On the contrary, if decision maker prefers dimension 1 more, then option 
$B$ will be eliminated, leaving both $A$ and $S_A$ to share the ``victory'', hence resulting in equation \ref{eq:similarityProbability}.

@@

MDFT has been confirmed to account for similarity, compromise, and attraction effects in multialternative choice scenarios \citep{roe2001multialternative}. It has previously been tested against such random utility models of choice as Logit and Probit and has been concluded to be a better fit to empirical data \citep{berkowitsch2014rigorously}. 
MDFT has been further adapted to account for preference shifts \citep{mohr2017attraction} and decision making under time restrictions \citep{diederich2003mdft}.

@@

According to MDbS, when comparing two tickets between Paris and New-York, the price of a ticket would be compared to the prices of other tickets in the choice set and also to the ones and which an individual has previously seen, but are not in the current choice set. Comparisons are ordinal, meaning that evidence accumulated towards the ``winner'' at a rate of one irregardless of how large the difference was.

@@

\begin{align}\label{probabilityOneIsFavored}
    P(m_{ij} \text{ is favored over } m_{kj}) = 
        \begin{cases}
        F(\beta_0 (| \frac{m_{ij} - m_{kj}}{m_{kj}} |- \beta_1)) & \text{if } A_i > X_i \\
        0 & \text{otherwise}
        \end{cases} ,
\end{align}
where $F$ is a logistic sigmoid function and $\beta_0$ and $\beta_1$ corresponde to the advantage value and the probability that this particular advantage value will be enough to be preferred. For example, consider the case where $\beta_0 = 0.1$ and $\beta_1 = 50$. This would mean that the advantage of 10\% would be preferred with 50\% probability. Consequently, if the difference is 20\%, then it will be preferred with 99\% probability. The logistic function brings the notion of ``soft'' comparison instead of ``hard'' comparison in which case the small differences would be ignored, while large differences would be extremely preferred \citep{noguchi2018multialternative}.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{staticFiles/contextEffectExampleScatterplot.png}
    \caption[MDbS’ account for big three context effects]{Example choice set to explain MDbS' account for big three context effects. $A$ and $B$ are considered original two alternatives (binary choice set). $D$ is dominated by $A$ on both dimensions. $C$ acts as compromise between the original two and $S$ is a similar option to $B$. Although these effects would be present in different variations of the choice set (for example one can make $A$ as compromise option), for the simplicity I will concentrate on this example.} % Add your description here
    \label{fig:MDBsContextExample} % This labels your figure for reference

@@

When adding option $D$ which is dominated by $A$ in both dimensions to binary choice set $A$ $B$ (hereafter binary choice set will be used instead of $A$ and $B$) one creates an attraction effect \citep{huberEtAl82, huberPuto83}. Huber et al. \citeyearonly{huberEtAl82} explained attraction effect via weight shifts for individuals. Addition of $D$ would make people to weigh dimension 2 more. Hence, $A$ and $D$ will have higher ``interest'' among people, and this is where $A$ wins over $D$.

@@

In MDbS this is explained by $\beta_0$ and $\beta_1$ from equation \ref{probabilityOneIsFavored}. Recall that, the reason of having the sigmoid function with arguments $\beta_0$ and $\beta_1$ is to make sure that small differences would be relatively ignored. This is in line with previous literature stating that people tend to ignore small differences between alternatives when making a choice \citep{kalwani1992consumer}. Hence, the small differences between $B$ and $S$ would be ignored, which would be translated to decreased probability of any of them winning over the other when being compared. This will indirectly increase the evidence accumulation for $A$ while resulting in ``shared'' evidence between $B$ and $S$.

@@

Because of MDbS nature of comparing dimensions with one another, for the purposes of this study, I can not use variables that do not differ across the menus. I can not utilize horizontal attributes in my analysis because they do not follow the standard ``greater the better'' mathematical approach to comparison either. However, these variables are used later in this dissertation in chapters \ref{chapter:jmrPaper} and \ref{chapter:hicssPaper}. Hence, I am bound to utilize solely four vertical attributes in my analysis. Table \ref{tab:descriptiveStats} provides descriptive information about variables. I have also multiplied all four vertical variables by \-1 to convert them to negative scale because of MDbS nature of comparing absolute values and the assumption that consumers prefer lower values along vertical dimensions.

@@

Choice set designers and engineers have long used ``Top n'' accuracy metrics when designing choice sets, or testing the performance of statistical models \citep{ricci2015recommender}. ``Top $n$'' accuracy metric posits whether or not, true class of the option matches the top $n$ predictions of the model. I will follow and adopt this metric because it is well established in the literature, mirrors the real-world decision making and it fits the contribution of the thesis the best. I will use ``Top 1'' accuracy which ranges from 0 to 1 as my metric of optimization for an individual choice set. Because choice sets in the data vary significantly in size, I will use average Top 1 accuracy metric weighted by menu size.This will ensure that smaller menus contribute proportional to their sizes. Also, to comply with differential evolution algorithm's aim of minimization, I will multiply this metric by -1. As an additional measure to explore the parameter space thoroughly, I employ multiple-run approach for differential evolution algorithm. Specifically, I will execute it ten times across the entire dataset. This reposition will allow me to further explore the parameter space, mitigating the risk of missing any region that can potentially contain optimal solution.
